Searching for standards of fairness in the transportation literature

### 2025-03-11 paper/paper.qmd
'paper' - submitted to journal.

---
title: "Searching for standards of fairness in the transportation literature"
csl: apa7.csl
author:
   - name: AAA #Anastasia Soukhov
    # affiliations:
    #   - ref: MAC
    # orcid: 0000-0003-4371-4831
    # email: soukho@mcmaster.ca
   - name: BBB #Ignacio Tiznado-Aitken
    # affiliations:
    #   - ref: TORONTO
    # orcid: 0000-0002-7385-2357
    # email: i.tiznadoaitken@utoronto.ca
   - name: CCC #Matthew Palm
    # affiliations:
    #   - ref: UNC
    # orcid: 0000-0002-8800-2777
    # email: palmmatt@unc.edu
   - name: DDD #Steven Farber
    # affiliations:
    #   - ref: TORONTO
    # orcid: 0000-0002-3870-5984
    # email: steven.farber@utoronto.ca
   - name: EEE #Antonio Páez
    # affiliations:
    #   - ref: MAC
    # orcid: 0000-0001-6912-9919
    # email: paezha@mcmaster.ca
# 
# affiliations:
#   - id: MAC
#     name: McMaster University
#     department: School of Earth, Environment and Society
#     city: Hamilton, ON
#     country: Canada
#   - id: TORONTO
#     name: University of Toronto Scarborough
#     department: Department of Geography and Planning
#     city: Toronto, ON
#     country: Canada
#   - id: UNC
#     name: The University of North Carolina at Chapel Hill
#     department: Department of City and Regional Planning
#     city: Chapel Hill, NC
#     country: USA
abstract: |
  This work provides a synthesis of how transportation fairness, justice and equity literature has operationalized standards. We first outline a flexible framework for engaging fairness questions and then apply it to collate the literature review's results. We find most articles were published within the last half decade and center urban areas, with 40% exploring Global South case studies. Income groups, followed by specific age groups (i.e., older adults) and people with disabilities, are most commonly the subjects of justice. Transit and pedestrian modes are the most frequently studied mobility tools; with many analyses being multimodal and comparative in nature. The benefits and burdens, the "What" of mobility, focuses on movement (i.e., quality of trips taken) or the potential for movement (i.e., accessibility). How the reviewed works conceptualize fairness is varied (e.g., Vertical Equity, Wellbeing, Rights-Based); but is often delineated with a supporting opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) standards are also prominent. While the reviewed literature is extensive, we conclude with calls to action for researchers, practitioners, and transportation advocates to drive meaningful change. These include developing standards based on rigorous conceptualizations, advancing systems-thinking approaches to fairness, ensuring data accessibility as a matter of justice, strengthening the connection between standards and lived experiences, and rigorously evaluating interventions and policies.

format: 
  #docx
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{graphicx}
        \usepackage{unicode-math}
        \usepackage{hyperref}
        \def\tightlist{}
        \usepackage{setspace}
        \doublespacing
        \usepackage{lineno}
        \linenumbers
    journal:
      name: Travel Behavior and Society
      formatting: preprint
      model: 3p
      cite-style: authoryear
keywords: [transportation; equity; justice; fairness; standards; transportation planning; literature review]
bibliography: bibliography.bib

#filters:
#  - docx-landscape.lua
---

@@

```{r, eval=FALSE}
df_all_sub_ob_just_reordered_plot
ggsave("figures/df_all_sub_ob_just_reordered_plot.tiff", dpi=300)
```

@@

```{r, eval=FALSE}
standards_conc_percentages_plot_and_table
ggsave("figures/standards_conc_percentages_plot_and_table.tiff", dpi=300)
```

@@

The primary research question and the protocol were initially defined by the authors, a group of experts in the field of transportation. The initial draft of the protocol was refined from preliminary searches of related reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021] and in consultation with research services librarians. The final protocol was released on OSF [@soukhovEquityJusticeConceptualizations2022], and the methods are summarised as follows.

@@

```{r plotting-who-what-where-when-plot-fig1}
#| label: fig-fig1
#| fig-cap: "The proportion of papers in each 'When', 'Where', 'Who', and 'What' by topics in each category. Topic categories were generated by the authors based on the reviewed literature."

knitr::include_graphics("figures/df_all_sub_ob_just_reordered_plot.tiff")
```
```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig2
#| fig-cap: "The proportion of papers by topics in the 'How' categories. 'How' is split into two categories: topics related to fairness conceptualisation (horizontal axis) and topics related to standards (vertical axis). Topic categories were generated by the authors based on the reviewed literature."

knitr::include_graphics("figures/standards_conc_percentages_plot_and_table.tiff")
```

@@

# Acknowledgements and declarations

@@

**Technology declaration:** During the preparation of this work, the first and second authors used ChatGPT to assist in reducing the word count in an early version of the “An appraisal of the lay of the land” section. After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.

@@

<!-- ## Why? A framework extension to analyse questions of justice {#sec-sect61} -->

<!-- To give an example drawing from the Canadian Charter of Rights and Freedoms, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex". Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person. Does this mean that they apply universally irrespective of age? Surely not, since no reasonable person would consider a toddler's demands for freedom of association as absolute. So, when do these rights fully apply in the life of a person? In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision. The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017]. Crook argued that the goal of teaching his kids (then aged 7, 8, 9 and 11) to ride transit to school was to raise self-reliant children [@stueckYoung2017]. While both sides discovered to everyone's surprise that no clear rules existed regarding children's independent transit use [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018]. If the public good is to reduce risk, shouldn't children be banned from riding in cars? The Canadian legal system concerns itself with fundamental freedoms and rights, one of which is the liberty to move freely. This provides an answer (of many) as to "why" we would even consider children as subjects of justice. But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail. Are children under the age of 10 subjects of justice (i.e., are they a "who") and who decides if they are (i.e., the courts or the parents in this case)? If so, is the ability to use transit unsupervised an object of justice that should be justly distributed (i.e., the "what") and under what circumstances (e.g, to school, to a social event; during the day, or late at night?). "When" (e.g., for all children going forward, for only 5 years) and "where" (e.g., only in the suburbs, across the country) do these questions apply? Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire and are encouraged to exercise independent mobility [@stueckCourt2020]. In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.  -->

<!-- Another perspective on the answer to "Why?" are cultural _norms_ that provide a rationale. In this case, the liberty of movement for all---including children---reigns supreme. But a competing concern is that the outside is dangerous, and it is so in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others. Parents may have over-pronounced 'stranger danger' based on neighbourhood and socio-economic characteristics of neighbourhoods [@francisLlBeDriving2017] and children themselves may fear potential violence based on real threats of victimization [@painParanoidParentingRematerializing2006]. Whose voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?".  -->

@@

```{r PRISMA-figA1}
#| label: fig-figA1
#| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies standards and EC signifies conceptualisations."
knitr::include_graphics("figures/Pflow_chart.tiff")
```

@@

```{r Search-query-A2}
#| label: fig-figA2
#| fig-cap: "The search query. TS = topic search (keywords, abstract, title). TASCA = subject categories. Green text area transportation system related terms, blue text are equity dimension related terms, purple text are equity/justice conceptualization related terms, and orange text are standards related terms. Hits corresponds the number of papers that the search yielded and was retained into the evidence selection process."
knitr::include_graphics("figures/Search-query.png")
```

@@

```{r methods_framework-fig1}
#| label: fig-figA3
#| fig-cap: "Evidence selection workflow. Step 1 (orange) is title and abstract screening, step 2 (green) is full-text review, and step 3 (purple) is data extraction."
knitr::include_graphics("figures/Methods_framework.png")
```

@@

:::landscape

## Select reviewed works summarized by the 5WH {#sec-sect63}

@@

dat <- data.frame(matrix(c("What?: **mobility and accessibilty**",
                           "Where? [@rivasHowAffordableTransportation2018] - South America (select cities)",

@@

                           "How?: The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.",
                           "What?: **mobility and accessibilty**",
                           "Where?: [@bharathyRevisitingClearFloor2018] - North America (USA - National)",

@@

                           "How?: The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards. Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",
                           "What?: **mobility and accessibilty**",

@@

                           "How?: Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.",
                           "What?: **mobility and accessibilty**",
                           "Where?: [@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)",

@@

                           "How?: Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minmum access): 1) Global: standard deviation (SD) distance to mean should be minimized. 2) Local: priority to minimise the differences with its neighbourhood",
                           "What?: **mobility and accessibilty**",
                           "Where?: [@zhengAreabasedEquitablePricing2020] - North America",
                           "This paper conceptualizies equity in the multimodal network (transit, car) being fair toll-pricing across differences in populatins value of time (VOT). VOT is determined based on household income, with lower income households having lower VOT and thus deserving of lower tolls (vertical equity). From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented.",
                           "How?: suggest that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.",

                           "What?: **environmental pollution**",
                           "Where?: [@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ",

@@

                           "How?: The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less. Residential locations should not be located within this distance threshold from the perspective of human health. **Environmental+** and **Population standards**: Uses the WHO NO² threshold as a point of comparison (annual concentrations of NO² should not exceed 40 μg/m-3). They argue that even through no neighbourhood, even those disproportionately low income, exceed the WHO limit in this case study, they still suggest that air pollution should not be disproportionately impacting disadvantaged neighbourhoods. It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)",
                           "What?: **environmental pollution**",
                           "Where?: [@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)",
                           "Geospatial analysis of naturally occurring boundaries in road-transport emissions and childrens respiratory health across a demographically diverse cityscape. Emperically identifies at what distance away from major roadways children are most impacted by transport-related pollution. This is framed in the perspective of children’s **well-being**. Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health.",
                           "How?: Finds that children (most vulnerable to air pollution - related to motoized traffic) are most impacted by air pollution within 283 m of a road way. This should be the distance threshold that schools and other childrens facilities are located.",
                           "What?: **health impacts**",
                           "Where? [@adlakhaMindGapGender2020] - Asia (Chennai, India)",

@@

                           "How?: All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.",
                           "What?: **health impacts**",
                           "Where? [@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)",
                           "The **well-being** of mothers, this paper examines the timely access to emergency obsteric and newborn care for child-bearing aged women in Uganda.",

@@

                           "What?: **health impacts**",
                           "Where?: [@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
                           "They use environmental pollution guidelines, but from the position of health. They investigate the impact of urban and transport planning on attributable mortality burden in Madrid and Barcelona and its distribution by socioeconomic status . Pre-mature mortality is linked to the exposure to pollution and motorized vehicles (**inequitable externalities**). These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).",
                           "How?: All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C. ",
                           "What?: **health impacts**",
                           "Where?: [@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)",

@@

                           "How?: perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",
                           "What?: **health impacts**",
                           "Where? [@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)",

@@

                           "How?: all households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this). Planners should prioritize socially disadvantaged areas to meeting these standards first.",

                           "What?: **transport-related safety**",
                           "Where?: [@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)",
                           "Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**.",
                           "How?: standards are suggested for both reactive and proactive analysis. First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better. No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract). Second, the lower the perceived safety, the better (i.e., if travel to school by ped. or bike is unsafe due to traffic conditions). No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract).",
                           "What?: **transport-related safety**",
                           "Where?: [@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima)",
                           "Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**.",
                           "How?: the study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width). The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.",

                           "What?: **mobility/accessibility and health impacts**",
                           "Where?: [@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**",

@@

                           "How?: 1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre. The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L",
                           "What?: **Mobility/ accessibility** and **health impacts**",
                           "Where?: [@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia)",

@@

                           "How: 1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school. 3) Access to health facility, percentage of households within 40 min walking time from a health facility. 4) Access to public transport, percentage of households within a distance of 500 m from a mini-bus stop. **Population standards**: 1) Adequate family income, percentage of households earning more than the official poverty line. 2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).",
                           "What?: **mobility/accessibility, health impacts, and safety**",
                           "Where? [@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",

@@

                           "How?: perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."

@@

```{r}
dat |>
  gt() |>
  tab_options(table.font.size = "xx-small") |>
  cols_label(X1 = "Dimension",
             X2 = "Continent",
             X3 = "Conceptualization",
             X4 = "Standard") |>
  tab_caption("Examples of Data Extraction")
```
:::

<!-- OLD ABSTRACT

>>>>

---
title: Searching for standards of fairness in the transportation literature
author:
  - name: AAA #Anastasia Soukhov
    #orcid: 0000-0003-4371-4831
    # email: soukhoa@example.com
    # affiliations: 
    #     - ref: MAC
    #       name: McMaster University
    #       department: School of Earth, Environment and Society
    #       address: 1280 Main Street West
    #       city: Hamilton, ON
    #       country: Canada
    #       postal-code: L8S 4L8
    # attributes:
    #     corresponding: true
  - name: BBB #Ignacio Tiznado-Aitken
    # orcid: 0000-0002-7385-2357
    # email: i.tiznadoaitken@utoronto.ca
    # affiliations:
    #     - ref: TORONTO
    #       name: University of Toronto Scarborough
    #       department: Department of Geography and Planning
    #       city: Toronto, ON
    #       country: Canada
    #       postal-code: M1C 1A4
  # - name: Matthew Palm
  #   # orcid: 0000-0002-8800-2777
  #   email: palmmatt@unc.edu
  #   affiliations:
  #       - ref: UNC
  #         name: The University of North Carolina at Chapel Hill
  #         department: Department of City and Regional Planning
  #         city: Chapel Hill, NC
  #         country: USA
  #         postal-code: 27599
  # - name: Steven Farber
  #   #orcid: 0000-0002-3870-5984
  #   email: steven.farber@utoronto.ca
  #   affiliations:
  #       - ref: TORONTO
  #         name: University of Toronto Scarborough
  #         department: Department of Geography and Planning
  #         city: Toronto, ON
  #         country: Canada
  #         postal-code: M1C 1A4
  # - name: Antonio Páez
  #   #orcid: 0000-0001-6912-9919
  #   email: paezha@mcmaster.ca
  #   affiliations: 
  #       - ref: MAC
  #         name: McMaster University
  #         department: School of Earth, Environment and Society
  #         address: 1280 Main Street West
  #         city: Hamilton, ON
  #         country: Canada
  #         postal-code: L8S 4L8
abstract: |
  This work provides a synthesis of how transportation fairness, justice and equity literature has operationalized standards. We first outline a flexible framework for engaging fairness questions and then apply it to collate the literature review's results. We find most articles were published within the last half decade and center urban areas, with 40% exploring Global South case studies. Income groups, followed by specific age groups (i.e., older adults) and people with disabilities, are most commonly the subjects of justice. Transit and pedestrian modes are the most frequently studied mobility tools; with many analyses being multimodal and comparative in nature. The benefits and burdens, the "What" of mobility, focuses on movement (i.e., quality of trips taken) or the potential for movement (i.e., accessibility). How the reviewed works conceptualize fairness is varied (e.g., Vertical Equity, Wellbeing, Rights-Based); but is often delineated with a supporting opportunity standard (37%) (i.e., number of parks within a 30-minute travel time) or population standard (36%) (i.e., bottom income quartile), though Infrastructure (i.e., level of service) and Environmental+ (i.e., air pollution) standards are also prominent. While the reviewed literature is extensive, we conclude with calls to action for researchers, practitioners, and transportation advocates to drive meaningful change. These include developing standards based on rigorous conceptualizations, advancing systems-thinking approaches to fairness, ensuring data accessibility as a matter of justice, strengthening the connection between standards and lived experiences, and rigorously evaluating interventions and policies.

keywords: [transportation; equity; justice; fairness; standards; transportation planning; literature review]
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    #keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
        \linenumbers
    journal:
      #name: Travel Behavior and Society
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

@@

```{r, eval=FALSE}
df_all_sub_ob_just_reordered_plot
ggsave("figures/df_all_sub_ob_just_reordered_plot.tiff", dpi=300) #fig1
```

@@

```{r, eval=FALSE}
standards_conc_percentages_plot_and_table
ggsave("figures/standards_conc_percentages_plot_and_table.tiff", dpi=300) #fig2
```

@@

The primary research question and the protocol were initially defined by the authors, a group of experts in the field of transportation. The initial draft of the protocol was refined from preliminary searches of related reviews [e.g., @zhangLiterature2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021] and in consultation with research services librarians. The final protocol was released on OSF [redacted for double-blind review] <!--[@soukhovEquityJusticeConceptualizations2022]-->, and the methods are summarised as follows.

@@

```{r plotting-who-what-where-when-plot-fig1}
#| label: fig-fig1
#| fig-cap: "The proportion of papers in each When, Where, Who, and What by topics in each category. Topic categories were generated by the authors based on the reviewed literature."

knitr::include_graphics("figures/Fig1.png")
```

```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig2
#| fig-cap: "The proportion of papers by topics in the How categories. How is split into two categories: topics related to fairness conceptualisation (horizontal axis) and topics related to standards (vertical axis). Topic categories were generated by the authors based on the reviewed literature."

knitr::include_graphics("figures/Fig2.png")
```

@@

<!--
# Acknowledgements and declarations

@@

**Technology declaration:** During the preparation of this work, the first and second authors used ChatGPT to assist in reducing the word count in an early version of the “Synthesis of Findings” section. After using this tool, all authors reviewed the manuscript and take full responsibility for the content of the publication.
-->

@@

@@

```{r PRISMA-figA1}
#| label: fig-figA1
#| fig-cap: "PRISMA flow diagram for the evidence selection process. ES signifies standards and EC signifies conceptualisations."
knitr::include_graphics("figures/Fig3_A.png")
```

@@

```{r Search-query-A2}
#| label: fig-figA2
#| fig-cap: "The search query. TS = topic search (keywords, abstract, title). TASCA = subject categories. Green text area transportation system related terms, blue text are equity dimension related terms, purple text are equity/justice conceptualization related terms, and orange text are standards related terms. Hits corresponds the number of papers that the search yielded and was retained into the evidence selection process."
knitr::include_graphics("figures/Fig4_A.png")
```

@@

```{r methods_framework-fig1}
#| label: fig-figA3
#| fig-cap: "Evidence selection workflow. Step 1 (orange) is title and abstract screening, step 2 (green) is full-text review, and step 3 (purple) is data extraction."
knitr::include_graphics("figures/Fig5_A.png")
```

@@

<!--:::landscape-->
\newpage

## Select reviewed works summarised using the 5WH {#sec-sect63}

@@

#1. What dimension?. 2) when and where?, 3) How is it conceptualised wrt who/what? 4. How is the standard set? 
dat <- data.frame(matrix(c("**Mobility and accessibility**",#What?: 
                           "[@rivasHowAffordableTransportation2018] - South America (select cities)", #Where? 

@@

                           " The financial burden of a basket of urban public transportation trips (60 trip fares, representing 30 round-trips per month) should not exceed 10% of household monthly income.", #How?:
                           "**Mobility and accessibility**", #What?: 
                           "[@bharathyRevisitingClearFloor2018] - North America (USA - National)", #Where?: 

@@

                           "The clear floor area for wheelchairs: 760 mm (30 in.) wide by 1220 mm (48 in.) in length as described by the ADA standards. Of note, this minimum clear floor area is insufficient for a variety of the WhMD users.",#How?: 
                           "**Mobility and accessibility**",#What?: 

@@

                           "Specifically for older populations (aged 65+), the following travel distances are suggested as equitable trip lengths to grocery stores per mode: Walking: less than or equal to 1500m, Combined transit and walking (less than or equal to 1000m (walking element)), Combined car and walking: less than or equal to 1000m less than or equal to 1000m (walking element)), Bicycle: less than or equal to 3000m in addition to travel time threshold of less than 15 mins.", #How?: 
                           "**Mobility and accessibility**",#What?: 
                           "[@wismadiSpatialPreferenceModelling2014] - Asia (Yogyakarta, Indonesia)", #Where?: 

@@

                           "Areas below the relative poverty line (of its neighbours) can only be located transport resources (i.e., measure in person*kms that can be travelled at car speed, i.e., mobility) based on the following 2 benchmarks (they can be considered, together as the floor/minimum access): 1) Global: standard deviation (SD) distance to mean should be minimized. 2) Local: priority to minimise the differences with its neighbourhood", #How?: 
                           "**Mobility and accessibility**", #What?: 
                           "[@zhengAreabasedEquitablePricing2020] - North America", #Where?: 
                           "This paper conceptualizes equity in the multimodal network (transit, car) being fair toll-pricing across differences in populations value of time (VOT). VOT is determined based on household income, with lower income households having lower VOT and thus deserving of lower tolls (vertical equity). From this perspective, a utilitarian perspective that seeks to minimize multimodal traffic congestion through introducing toll-pricing based on VOT is implemented.",
                           "Suggests that a toll-pricing scheme based on individuals travel value-of-time (lower income people have a lower VOT) is equitable.", #How?:

                           "**Environmental pollution**", #What?: 
                           "[@carrierApplicationThreeMethods2014] - North America (Montreal, Canada) ", #Where?:

@@

                           "The literature suggests that the health implications from the transport-related air pollution from major roadways is most acute at residential distance locations of 200 m or less. Residential locations should not be located within this distance threshold from the perspective of human health. **Environmental+** and **Population standards**: Uses the WHO NO² threshold as a point of comparison (annual concentrations of NO² should not exceed 40 μg/m-3). They argue that even through no neighbourhood, even those disproportionately low income, exceed the WHO limit in this case study, they still suggest that air pollution should not be disproportionately impacting disadvantaged neighbourhoods. It can be interpreted that they use the WHO threshold as a minimum threshold and suggest that air pollution levels should not be impacting disadvantaged populations disproportionately ( a relative population standard)", #How?: 
                           "**Environmental pollution**", #what?
                           "[@jephcoteGeospatialAnalysisNaturally2013] - Europe (Leicester, UK)", #Where?: 
                           "Geospatial analysis of naturally occurring boundaries in road-transport emissions and children's respiratory health across a demographically diverse cityscape. Emperically identifies at what distance away from major roadways children are most impacted by transport-related pollution. This is framed in the perspective of children’s **well-being**. Children are at most risk for acute respiratory distress from elevated levels of air pollution, and as such planning should consider this point of public health.",
                           "Finds that children (most vulnerable to air pollution - related to motorized traffic) are most impacted by air pollution within 283 m of a road way. This should be the distance threshold that schools and other children's facilities are located.", #How?: 
                           "**Health impacts**", #What?: 
                           "[@adlakhaMindGapGender2020] - Asia (Chennai, India)", #Where? 

@@

                           "All people should get 150 min of moderate activity a week or 75 min of vigorous physical activity per week.", #How?: 
                           "**Health impacts**", #How?: 
                           "[@savingmothersDidSavingMothers2019] - Africa (Select urban and rural regions in Uganda)", #Where? 
                           "The **well-being** of mothers, this paper examines the timely access to emergency obstetric and newborn care for child-bearing aged women in Uganda.",

@@

                           "**Health impacts**", #What?: 
                           "[@iungmanImpactUrbanTransport2021] - Europe (Madrid and Barcelona, Spain)",
                           "They use environmental pollution guidelines, but from the position of health. They investigate the impact of urban and transport planning on attributable mortality burden in Madrid and Barcelona and its distribution by socioeconomic status . Pre-mature mortality is linked to the exposure to pollution and motorized vehicles (**inequitable externalities**). These externalities should not be impacting people disproportionately (**vertical equity**) and should be even across space (**spatial equity**).", #Where?: 
                           "All minimum thresholds, if exceeded this is inequitable: NO² concentration 40 μg/m³; PM 2.5 concentration 10 μg/m³; Noise 53dB for average 24 hours; Living with 300 m crow-flies distance from at least .5 hectares of greenspace; and a Change of air temperature of at least 1 ⁰C. ", #How?: 
                           "**Health impacts**", #What?: 
                           "[@mehdizadehWalkingTimeSchool2017] - Asia (Rasht, Iran)", #Where?: 

@@

                           "Perceived walking time to school for students aged 7-9 yrs is 10 mins, and the longer the PWTS the less likely they were to use an active mode to travel to school.",#How?: 
                           "**Health impacts**", #What?: 
                           "[@murphySupermarketAccessTransport2017] - Oceania (Melbourne, Australia)", #Where? 

@@

                           "All households should be sufficiently active (greater than 150 min and at least 5 sessions) and households should be within 1 km euclidean distance to supermarket (80-90% of the dwellings should meet this). Planners should prioritize socially disadvantaged areas to meeting these standards first.",#How?:

                           "**Transport-related safety**", #What?: 
                           "[@ferenchakEquityAnalysisProactively2019] - North America (Denver, USA)", #Where?: 
                           "Operationalizes and compares an equity analysis of proactively- and reactively-identified traffic safety issues from the perspective of **Spatial equity**, **Vertical equity** and **Inequitable exposure to externalities**.",
                           "Standards are suggested for both reactive and proactive analysis. First, the lower the number of collisions on the road with pedestrians/cyclists (i.e., reactive safety analysis), the better. No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract). Second, the lower the perceived safety, the better (i.e., if travel to school by ped. or bike is unsafe due to traffic conditions). No/minimal inequalities for general population vs. equity seeking groups (high proportion of POC and/or low income in tract).",#How?:
                           "**Transport-related safety**", #What?: 
                           "[@zheEvaluationSharedUse2008] - Asia (Tokyo, Takamatsu, and Tokushima)", #Where?: 
                           "Evaluates the observed safety of  shared use pedestrian and bicycle paths from the perspective of **well-being**.",
                           "The study suggests that the safety threshold for bicycles and pedestrians to coexist on shared infrastructure is less than 0.5 pedestrians/minute per metre of sidewalk (width) and less than 3.0 cyclists/minute per metre of sidewalk (width). The standard for pedestrian/bicycle share use in terms of hourly traffic volume is less than 26 pedestrians / hour and 108 cyclists / hour for 2m wide sidewalks.", #How?:

                           "**Mobility, accessibility and health impacts**", #What?: 
                           "[@aldertonWhatMeaningUrban2019] - Asia (Bangkok, Thailand) – **Mobility/ accessibility** and **health**", #Where?: 

@@

                           "1) Green space: % of residents living < 400 m from public open space, a large park (> 1.5ha), and/or local park, 2) transit access: % of residents living < 400 m of a local bus stop and <800 m of train station, 3) Facilities: % of residents living < 400 m of a community centre. The following **Infrastructure standard** is suggested: Canal water quality - dissolved oxygen content of equal to or less than 2.0 mL/L", #How?: 
                           "**Mobility/ accessibility** and **health impacts**", #What?: 
                           "[@berheAdaptationDissonanceQuality2014] - Africa (Mekelle, Ethiopia)", #Where?: 

@@

                           "1 & 2) Access to primary or secondary education facility, percentage of households living within 1 km or 2km (walking distance), respectively from a primary school or secondary school. 3) Access to health facility, percentage of households within 40 min walking time from a health facility. 4) Access to public transport, percentage of households within a distance of 500 m from a mini-bus stop. **Population standards**: 1) Adequate family income, percentage of households earning more than the official poverty line. 2) Subjective QoL is constructed based on the households level of satisfaction for each of the eight indicators using a six point Likert-scale (1=very satisfied to 6=very dissatisfied).", #How: 
                           "**Mobility, accessibility, health impacts, and safety**", #What?: 
                           "[@agostfelipInclusiveModelAssessing2021] - Europe (Castellon, Spain)",#Where? 

@@

                           "Perceived pedestrian perception on protection, ease, equitable access, mobility, identity, and enjoyment must be met."#How?: 

@@

# dat |>
#   gt() |>
#   tab_options(table.font.size = "xx-small") |>
#   cols_label(X1 = "Dimension (What?)",
#              X2 = "Continent (Where?),
#              X3 = "Conceptualization",
#              X4 = "Standard") |>
#   tab_caption("Examples of Data Extraction") |>
#   fmt_markdown(columns = everything()) |>
#   tab_options(table.width = px(400))

colnames(dat) <- c("What dimensions?", "When and where?", "How is fairness conceptualised wrt Who and What?","How is the standard defined?")
```

```{r}
#| label: tbl-tabA63
#tbl-cap: "Sample extracted articles."
ft <- dat |>
flextable() |>
colformat_md() |>
width(j = c(1,2), width = 2, unit = "cm")|>
width(j = c(3,4), width = 6, unit = "cm") |>
fontsize(size = 8, part = "all")
ft
```

<!-- :::-->
